---
title: "NHST vs estimation"
author: "Ian Hussey"
output:
  html_document:
    code_folding: hide
    highlight: haddock
    theme: flatly
    toc: yes
    toc_float: yes
---

# Simulated data where effect is absent

Simulated data with Cohen's d = 0

```{r echo=FALSE, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}

sequential_analysis <- function(cohens_d, participants){
  
  require(simstudy)
  require(tidyverse)
  
  parameters_a <- defData(varname = "Score", dist = "normal", formula = cohens_d, variance = 1, id = "idnum")
  parameters_b <- defData(varname = "Score", dist = "normal", formula = 0,   variance = 1, id = "idnum")
  
  # generate required number of data points using above parameters
  # for random participant numbers/order
  data_participant_ids <- genData(participants, parameters_a) %>%
    rename(ordering = Score)
  
  # for condition A (mu = 0)
  data_a <- genData(participants/2, parameters_a) %>%
    mutate(Condition = "A")
  
  # for condition B (mu = cohens_d)
  data_b <- genData(participants/2, parameters_b) %>%
    mutate(Condition = "B",
           idnum = idnum + participants/2) 
  
  data <- rbind(data_a, data_b) %>%
    mutate(SCIAT_D1 = round(Score, 2)) %>%
    arrange(idnum, Condition) %>%
    left_join(data_participant_ids, by = "idnum") %>%
    arrange(ordering) %>%
    rownames_to_column(var = "participant_n") %>%
    select(-idnum, -ordering, -Score) %>%
    rename(IAT_condition = Condition) %>%
    mutate(participant_n = as.numeric(participant_n))
  
  
  # dependencies
  library(effsize)
  library(tidyverse)
  
  # sequentially tested p values
  p_seq <- function(data){
    output <- NULL
    for (i in 10:nrow(data)) {
      p_seq <- data %>%
        filter(between(participant_n, 1, i)) %>%
        do(p = t.test(SCIAT_D1~IAT_condition, 
                      data = ., 
                      var.equal=TRUE, 
                      paired=FALSE)$p.value)
      output[i] <- p_seq$p %>% as.numeric()
    }
    output <- output %>% as.data.frame()
    colnames(output) <- "p"
    return(output)
  } 
  
  p_seq <- p_seq(data) %>%
    rownames_to_column(var = "participant_n") %>%
    mutate(participant_n = as.numeric(participant_n))
  
  # sequentially tested cohens d estimates
  # lower 95% CIs
  d_lwr_seq <- function(data){
    output <- NULL
    for (i in 10:nrow(data)) {
      d_lwr <- data %>%
        filter(between(participant_n, 1, i)) %>%
        do(d = cohen.d(SCIAT_D1~IAT_condition, 
                       data = ., 
                       var.equal=TRUE, 
                       paired=FALSE)$conf.int[1])
      output[i] <- d_lwr$d %>% as.numeric()*-1
    }
    output <- output %>% as.data.frame()
    colnames(output) <- "lwr"
    return(output)
  } 
  
  lwr <- d_lwr_seq(data) %>%
    rownames_to_column(var = "participant_n") %>%
    mutate(participant_n = as.numeric(participant_n), 
           lwr = lwr*-1)
  
  # upper 95% CIs
  d_upr_seq <- function(data){
    output  <- NULL
    for (i in 10:nrow(data)) {
      d_upr <- data %>%
        filter(between(participant_n, 1, i)) %>%
        do(d = cohen.d(SCIAT_D1~IAT_condition, 
                       data = ., 
                       var.equal=TRUE, 
                       paired=FALSE)$conf.int[2])
      output[i] <- d_upr$d %>% as.numeric()*-1
    }
    output <- output %>% as.data.frame()
    colnames(output) <- "upr"
    return(output)
  } 
  
  upr <- d_upr_seq(data) %>%
    rownames_to_column(var = "participant_n") %>%
    mutate(participant_n = as.numeric(participant_n),
           upr = upr*-1)
  
  # estimates
  d_est_seq <- function(data){
    output <- NULL
    for (i in 10:nrow(data)) {
      d_est <- data %>%
        filter(between(participant_n, 1, i)) %>%
        do(d = cohen.d(SCIAT_D1~IAT_condition, 
                       data = ., 
                       var.equal=TRUE, 
                       paired=FALSE)$estimate)
      output[i] <- d_est$d %>% as.numeric()*-1
    }
    output <- output %>% as.data.frame()
    colnames(output) <- "est"
    return(output)
  } 
  
  est <- d_est_seq(data) %>%
    rownames_to_column(var = "participant_n") %>%
    mutate(participant_n = as.numeric(participant_n),
           est = est*-1)
  
  # combine data frames
  full_data <- data %>%
    left_join(p_seq, by = "participant_n") %>%
    full_join(upr, by = "participant_n") %>%
    full_join(est, by = "participant_n") %>%
    full_join(lwr, by = "participant_n") %>%
    filter(participant_n > 10) # drop first 10 participants as these can't produces estimates/p values

  return(full_data)
}

```
  
## Simulation 1

Sequentially tested *p* values
  
```{r echo=FALSE, fig.height=3, fig.width=5, message=FALSE, warning=FALSE}
  
set.seed(423) 

full_data <- sequential_analysis(cohens_d = 0, participants = 200)

# sequential tested p values
ggplot(full_data) +
  geom_line(aes(x = participant_n, y = p)) + # , color = (p < 0.05), group = 1
  geom_hline(yintercept = 0.05, linetype = "dashed") +
  ylab("p value") +
  xlab("Participants")

```

Sequentially tested Cohen's d estimates

```{r echo=FALSE, fig.height=3, fig.width=5, message=FALSE, warning=FALSE}

# sequential estimated d values
ggplot(full_data) +
  geom_ribbon(aes(x = participant_n, ymin = lwr, ymax = upr), fill = "black", alpha = .2) + # grey70 # , color = (p < 0.05), group = 1
  geom_line(aes(participant_n, est)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  ylab("Cohen's d") +
  xlab("Simulated participants")

```

## Simulation 2

Sequentially tested *p* values
  
```{r echo=FALSE, fig.height=3, fig.width=5, message=FALSE, warning=FALSE}
  
set.seed(12345) 

full_data <- sequential_analysis(cohens_d = 0, participants = 200)

# sequential tested p values
ggplot(full_data) +
  geom_line(aes(x = participant_n, y = p)) + # , color = (p < 0.05), group = 1
  geom_hline(yintercept = 0.05, linetype = "dashed") +
  ylab("p value") +
  xlab("Participants")

```

Sequentially tested Cohen's *d* estimates

```{r echo=FALSE, fig.height=3, fig.width=5, message=FALSE, warning=FALSE}

# sequential estimated d values
ggplot(full_data) +
  geom_ribbon(aes(x = participant_n, ymin = lwr, ymax = upr), fill = "black", alpha = .2) + # grey70 # , color = (p < 0.05), group = 1
  geom_line(aes(participant_n, est)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  ylab("Cohen's d") +
  xlab("Simulated participants")

```

## Simulation 3

Sequentially tested *p* values
  
```{r echo=FALSE, fig.height=3, fig.width=5, message=FALSE, warning=FALSE}
  
set.seed(42)  

full_data <- sequential_analysis(cohens_d = 0, participants = 200)

# sequential tested p values
ggplot(full_data) +
  geom_line(aes(x = participant_n, y = p)) + # , color = (p < 0.05), group = 1
  geom_hline(yintercept = 0.05, linetype = "dashed") +
  ylab("p value") +
  xlab("Participants")

```

Sequentially tested Cohen's *d* estimates

```{r echo=FALSE, fig.height=3, fig.width=5, message=FALSE, warning=FALSE}

# sequential estimated d values
ggplot(full_data) +
  geom_ribbon(aes(x = participant_n, ymin = lwr, ymax = upr), fill = "black", alpha = .2) + # grey70 # , color = (p < 0.05), group = 1
  geom_line(aes(participant_n, est)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  ylab("Cohen's d") +
  xlab("Simulated participants")

```

## Simulation 4

Sequentially tested *p* values
  
```{r echo=FALSE, fig.height=3, fig.width=5, message=FALSE, warning=FALSE}
  
set.seed(198723) 

full_data <- sequential_analysis(cohens_d = 0, participants = 200)

# sequential tested p values
ggplot(full_data) +
  geom_line(aes(x = participant_n, y = p)) + # , color = (p < 0.05), group = 1
  geom_hline(yintercept = 0.05, linetype = "dashed") +
  ylab("p value") +
  xlab("Participants")

```

Sequentially tested Cohen's *d* estimates

```{r echo=FALSE, fig.height=3, fig.width=5, message=FALSE, warning=FALSE}

# sequential estimated d values
ggplot(full_data) +
  geom_ribbon(aes(x = participant_n, ymin = lwr, ymax = upr), fill = "black", alpha = .2) + # grey70 # , color = (p < 0.05), group = 1
  geom_line(aes(participant_n, est)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  ylab("Cohen's d") +
  xlab("Simulated participants")

```

# Simulated data where effect is present

Simulated data with Cohen's d = 0.5

## Simulation 1

Sequentially tested *p* values
  
```{r echo=FALSE, fig.height=3, fig.width=5, message=FALSE, warning=FALSE}
  
set.seed(423) # initially volitile, but sigificant roughly where power analysis suggests it should be (c. n = 128)

full_data <- sequential_analysis(cohens_d = 0.5, participants = 200)

# sequential tested p values
ggplot(full_data) +
  geom_line(aes(x = participant_n, y = p)) + # , color = (p < 0.05), group = 1
  geom_hline(yintercept = 0.05, linetype = "dashed") +
  ylab("p value") +
  xlab("Participants")

```

Sequentially tested Cohen's d estimates

```{r echo=FALSE, fig.height=3, fig.width=5, message=FALSE, warning=FALSE}

# sequential estimated d values
ggplot(full_data) +
  geom_ribbon(aes(x = participant_n, ymin = lwr, ymax = upr), fill = "black", alpha = .2) + # grey70 # , color = (p < 0.05), group = 1
  geom_line(aes(participant_n, est)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  ylab("Cohen's d") +
  xlab("Simulated participants")

```

## Simulation 2

Sequentially tested *p* values
  
```{r echo=FALSE, fig.height=3, fig.width=5, message=FALSE, warning=FALSE}
  
set.seed(12345) # initially sigificant but not at n = 200

full_data <- sequential_analysis(cohens_d = 0.5, participants = 200)

# sequential tested p values
ggplot(full_data) +
  geom_line(aes(x = participant_n, y = p)) + # , color = (p < 0.05), group = 1
  geom_hline(yintercept = 0.05, linetype = "dashed") +
  ylab("p value") +
  xlab("Participants")

```

Sequentially tested Cohen's *d* estimates

```{r echo=FALSE, fig.height=3, fig.width=5, message=FALSE, warning=FALSE}

# sequential estimated d values
ggplot(full_data) +
  geom_ribbon(aes(x = participant_n, ymin = lwr, ymax = upr), fill = "black", alpha = .2) + # grey70 # , color = (p < 0.05), group = 1
  geom_line(aes(participant_n, est)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  ylab("Cohen's d") +
  xlab("Simulated participants")

```

## Simulation 3

Sequentially tested *p* values
  
```{r echo=FALSE, fig.height=3, fig.width=5, message=FALSE, warning=FALSE}
  
set.seed(42)  # significant and quickly

full_data <- sequential_analysis(cohens_d = 0.5, participants = 200)

# sequential tested p values
ggplot(full_data) +
  geom_line(aes(x = participant_n, y = p)) + # , color = (p < 0.05), group = 1
  geom_hline(yintercept = 0.05, linetype = "dashed") +
  ylab("p value") +
  xlab("Participants")

```

Sequentially tested Cohen's *d* estimates

```{r echo=FALSE, fig.height=3, fig.width=5, message=FALSE, warning=FALSE}

# sequential estimated d values
ggplot(full_data) +
  geom_ribbon(aes(x = participant_n, ymin = lwr, ymax = upr), fill = "black", alpha = .2) + # grey70 # , color = (p < 0.05), group = 1
  geom_line(aes(participant_n, est)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  ylab("Cohen's d") +
  xlab("Simulated participants")

```

## Simulation 4

Sequentially tested *p* values
  
```{r echo=FALSE, fig.height=3, fig.width=5, message=FALSE, warning=FALSE}
  
set.seed(198723)  # significant and quickly

full_data <- sequential_analysis(cohens_d = 0.5, participants = 200)

# sequential tested p values
ggplot(full_data) +
  geom_line(aes(x = participant_n, y = p)) + # , color = (p < 0.05), group = 1
  geom_hline(yintercept = 0.05, linetype = "dashed") +
  ylab("p value") +
  xlab("Participants")

```

Sequentially tested Cohen's *d* estimates

```{r echo=FALSE, fig.height=3, fig.width=5, message=FALSE, warning=FALSE}

# sequential estimated d values
ggplot(full_data) +
  geom_ribbon(aes(x = participant_n, ymin = lwr, ymax = upr), fill = "black", alpha = .2) + # grey70 # , color = (p < 0.05), group = 1
  geom_line(aes(participant_n, est)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  ylab("Cohen's d") +
  xlab("Simulated participants")

```

# Real data

Data taken from an actual experiment examining the influence of an evaluative learning task involving valenced stimuli and neutral and unknown Chinese characters. Between conditions, a source stimulus was varied to be either positively or negatively valenced. Outcome task was automatic evaluations of the Chinese characters on a Single-Category IAT. 

```{r echo=FALSE, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}

# dependencies
library(effsize)
library(tidyverse)

# get data
## analogical learning via IAT experiment 3, opposite attributes conditions only 
data <- read.csv("data.csv")

# sequentially tested p values
p_seq <- function(data){
  output <- NULL
  for (i in 10:nrow(data)) {
    p_seq <- data %>%
      filter(between(participant_n, 1, i)) %>%
      do(p = t.test(SCIAT_D1~IAT_condition, 
                    data = ., 
                    var.equal=TRUE, 
                    paired=FALSE)$p.value)
    output[i] <- p_seq$p %>% as.numeric()
  }
  output <- output %>% as.data.frame()
  colnames(output) <- "p"
  return(output)
} 

p_seq <- p_seq(data) %>%
  rownames_to_column(var = "participant_n") %>%
  mutate(participant_n = as.numeric(participant_n))


# sequentially tested cohens d estimates
# lower 95% CIs
d_lwr_seq <- function(data){
  output <- NULL
  for (i in 10:nrow(data)) {
    d_lwr <- data %>%
      filter(between(participant_n, 1, i)) %>%
      do(d = cohen.d(SCIAT_D1~IAT_condition, 
                     data = ., 
                     var.equal=TRUE, 
                     paired=FALSE)$conf.int[1])
    output[i] <- d_lwr$d %>% as.numeric()*-1
  }
  output <- output %>% as.data.frame()
  colnames(output) <- "lwr"
  return(output)
} 

lwr <- d_lwr_seq(data) %>%
  rownames_to_column(var = "participant_n") %>%
  mutate(participant_n = as.numeric(participant_n), 
         lwr = lwr*-1)

# upper 95% CIs
d_upr_seq <- function(data){
  output  <- NULL
  for (i in 10:nrow(data)) {
    d_upr <- data %>%
      filter(between(participant_n, 1, i)) %>%
      do(d = cohen.d(SCIAT_D1~IAT_condition, 
                     data = ., 
                     var.equal=TRUE, 
                     paired=FALSE)$conf.int[2])
    output[i] <- d_upr$d %>% as.numeric()*-1
  }
  output <- output %>% as.data.frame()
  colnames(output) <- "upr"
  return(output)
} 

upr <- d_upr_seq(data) %>%
  rownames_to_column(var = "participant_n") %>%
  mutate(participant_n = as.numeric(participant_n),
         upr = upr*-1)

# estimates
d_est_seq <- function(data){
  output <- NULL
  for (i in 10:nrow(data)) {
    d_est <- data %>%
      filter(between(participant_n, 1, i)) %>%
      do(d = cohen.d(SCIAT_D1~IAT_condition, 
                     data = ., 
                     var.equal=TRUE, 
                     paired=FALSE)$estimate)
    output[i] <- d_est$d %>% as.numeric()*-1
  }
  output <- output %>% as.data.frame()
  colnames(output) <- "est"
  return(output)
} 

est <- d_est_seq(data) %>%
  rownames_to_column(var = "participant_n") %>%
  mutate(participant_n = as.numeric(participant_n),
         est = est*-1)

# combine data frames
full_data <- data %>%
  left_join(p_seq, by = "participant_n") %>%
  full_join(upr, by = "participant_n") %>%
  full_join(est, by = "participant_n") %>%
  full_join(lwr, by = "participant_n") %>%
  filter(participant_n > 10) # drop first 10 participants as these can't produces estimates/p values

```

Sequentially tested *p* values

```{r echo=FALSE, fig.height=3, fig.width=5, message=FALSE, warning=FALSE}

# sequential tested p values
ggplot(full_data) +
  geom_line(aes(x = participant_n, y = p)) + # , color = (p < 0.05), group = 1
  geom_hline(yintercept = 0.05, linetype = "dashed") +
  ylab("p value") +
  xlab("Participants")

```

Sequentially tested Cohen's *d* estimates

```{r echo=FALSE, fig.height=3, fig.width=5, message=FALSE, warning=FALSE}

# sequential estimated d values
ggplot(full_data) +
  geom_ribbon(aes(x = participant_n, ymin = lwr, ymax = upr), fill = "black", alpha = .2) + # grey70 # , color = (p < 0.05), group = 1
  geom_line(aes(participant_n, est)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  ylab("Cohen's d") +
  xlab("Participants")

```