---
title: "Exploring self-reported beliefs about age"
author: "Ian Hussey^[Ghent University. Email: ian.hussey@ugent.be]"
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: This script processes data from the Attitudes 2.0 study from raw data to
  an analysable format. It includes all necessary corrections, scoring, and organisation
  for full reproducability.
output:
  html_document:
    code_folding: hide
    highlight: haddock
    number_sections: yes
    theme: flatly
    toc: yes
    toc_float: yes
---

licenses: 
  code: GPLv3+
  data: CC0

```{r get and clean data}

# dependencies ------------------------------------------------------------

library(foreign)
library(tidyverse)
library(psych)

# get data ----------------------------------------------------------------

# data from project implicit demo datasets for Age IAT 2015 data https://osf.io/pbzcu/

# convert spss to rdata
#df <- read.spss("Age IAT.public.2015.sav", to.data.frame=TRUE)

# save as rdata for future reference
#save(df, file = "Age IAT.public.2015.rdata")

# load rdata file
load(file = "data/Age IAT.public.2015.rdata")

# tidy data ---------------------------------------------------------------

trimmed_df <-
  df %>%
  select(session_id,
         age,
         sex,
         choosetobe,
         feel,
         hopetolive,
         othersthink,
         atoma,  # A person moves from being an adult to middle-aged at what age?
         ctoya,  # A person moves from being a child to being a young adult at what age?
         matoo,  # A person moves from being middle-aged to being old at what age?
         yatoa) %>%  # A person moves from being a young adult to being an adult at what age?
  filter(!is.na(age)) %>%
  filter(sex == "m" | sex == "f")

choosetobe_data <-
  trimmed_df %>%
  filter(!is.na(choosetobe)) %>%
  select(age, choosetobe, sex) %>%
  mutate(age = as.integer(age),
         choosetobe = as.integer(choosetobe))
```

# Research question
"How does age related to the age people would choose to be?"

Descriptive stats

```{r choose to be - descriptives}
descriptives <- describe(choosetobe_data) %>% select(n, mean, sd)
descriptives
```

## Plots
### Age distribution

```{r }
mean_age <- descriptives["age","mean"]

ggplot(data = choosetobe_data) +
  geom_density(aes(x = age), adjust = 4, colour = "red", fill = "red", alpha = 0.2) +
  geom_vline(xintercept = mean_age, colour = "red") +
  theme_minimal()
```

### Preferred age distribution

```{r}
mean_agechoosetobe <- descriptives["choosetobe","mean"]

ggplot(data = choosetobe_data) +
  geom_density(aes(x = choosetobe), adjust = 4, colour = "blue", fill = "blue", alpha = 0.2) +
  geom_vline(xintercept = mean_agechoosetobe, colour = "blue") +
  theme_minimal()
```

Choose to be split by gender

```{r }
ggplot(data = choosetobe_data) +
  geom_density(aes(x = choosetobe, group = sex, fill = sex, colour = sex), adjust = 4, alpha = 0.2) +
  geom_vline(xintercept = mean_age, colour = "red") +
  theme_minimal()
```

### Comparison

```{r}
ggplot(data = choosetobe_data) +
  geom_density(aes(x = age), adjust = 4, colour = "red", fill = "red", alpha = 0.2) +
  geom_vline(xintercept = mean_age, colour = "red") +
  geom_density(aes(x = choosetobe), adjust = 4, colour = "blue", fill = "blue", alpha = 0.2) +
  geom_vline(xintercept = mean_agechoosetobe, colour = "blue") +
  theme_minimal()
```

## Comparing distributions using t tests

```{r}
attach(choosetobe_data)

t.test(age, choosetobe, 
       alternative = "two.sided",
       paired = TRUE)
```

## Comparing 

### Plot raw data using scatterplot

```{r}
ggplot(data = choosetobe_data) +
  geom_point(aes(x = age, y = choosetobe), colour = "darkgreen", alpha = 0.1) +
  theme_minimal()
```

```{r}
ggplot(data = choosetobe_data) +
  geom_point(aes(x = age, y = choosetobe), colour = "darkgreen", alpha = 0.1) +
  geom_smooth(aes(x = age, y = choosetobe), method = "lm", colour = "darkorange") +
  theme_minimal()
```


```{r}
require(schoRsch)

choosetobe_trimmed_outlier_df <-
  choosetobe_data %>%
  schoRsch::outlier(dv = "age", 
                    res.name = "age_outlier",
                    todo = "nothing", 
                    upper.z = 2.5, 
                    lower.z = -2.5) %>%
  schoRsch::outlier(dv = "choosetobe",                   # not producing consisent results!
                    res.name = "choosetobe_outlier",
                    todo = "nothing", 
                    upper.z = 2.5, 
                    lower.z = -2.5) %>%  
  mutate(exclude = age_outlier + choosetobe_outlier) %>%
  filter(exclude == 0)

ggplot(data = choosetobe_trimmed_outlier_df) +
  geom_point(aes(x = age, y = choosetobe, group = sex, fill = sex, colour = sex), alpha = 0.1) +
  geom_smooth(aes(x = age, y = choosetobe, group = sex, colour = sex), method = "lm") +
  theme_minimal()
```


```{r}
# linear regression
model_1 <- lm(choosetobe ~ age,
              data = choosetobe_trimmed_outlier_df)
summary(model_1)
```
half their age plus 10

nb it is equally correct to say that participants want to be 2 years younger than they are on average, and that participants want to be half their age plus 10. 



# other analyses
## age ~ age others think

```{r}
ggplot(trimmed_df, aes(x = age, y = othersthink)) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = "lm") +
  theme_minimal()
```

## systematic irrational belief about what age we look

```{r}
model_2 <- lm(data = trimmed_df,
              formula = othersthink ~ age)
summary(model_2)
```

## distribution of beliefs about age transitions

```{r}
# reshape age transitions data
transitions <- 
  trimmed_df %>%
  rename(`adult to middle` = atoma,  # A person moves from being an adult to middle-aged at what age?
         `child to ya` = ctoya,  # A person moves from being a child to being a young adult at what age?
         `middle to old` = matoo,  # A person moves from being middle-aged to being old at what age?
         `ya to adult` = yatoa) %>%
  gather(transition_type, transition_age, c(`adult to middle`, `child to ya`, `middle to old`, `ya to adult`)) %>%
  mutate(transition_type = as.factor(transition_type)) %>%
  filter(!is.na(transition_age), !is.na(transition_type))
```



```{r}
ggplot(transitions, aes(x = transition_age, colour = transition_type, fill = transition_type)) +
  geom_density(alpha = 0.5, adjust = 5) +  # try changing the kernal smoothing width via adjust.
  theme_minimal()
```

## consensus at younger ages, disagreement at older ages

```{r}
model_1 <- lm(data = transitions,
              formula = transition_age ~ transition_type)
drop1(model_1, scope = .~., test = "Chisq")
summary(model_1)
```

## relationship between age and beliefs about age transitions

```{r}
ggplot(transitions, aes(x = age, y = transition_age, colour = transition_type, fill = transition_type)) +
  geom_smooth(method = "lm") +
  theme_minimal()
```

## beliefs about age transitions are associated with current age 

```{r}
model_2 <- lm(data = transitions,
              formula = transition_age ~ age * transition_type)
drop1(model_2, scope = .~., test = "Chisq")
summary(model_2)
```


