---
title: "Exploring self-reported beliefs about age"
subtitle: "Data cleaning and processing"
author: "Ian Hussey^[Ghent University. Email: ian.hussey@ugent.be]"
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: This script processes data from the Attitudes 2.0 study from raw data to
  an analysable format. It includes all necessary corrections, scoring, and organisation
  for full reproducability.
output:
  html_document:
    code_folding: hide
    highlight: haddock
    number_sections: yes
    theme: flatly
    toc: yes
    toc_float: yes
---

```{r get and clean data}

# dependencies ------------------------------------------------------------

library(foreign)
library(tidyverse)
library(psych)

# get data ----------------------------------------------------------------

# data from project implicit demo datasets for Age IAT 2015 data https://osf.io/pbzcu/

# convert spss to rdata
#df <- read.spss("Age IAT.public.2015.sav", to.data.frame=TRUE)

# tidy data ---------------------------------------------------------------

# trimmed_df <-
#   df %>%
#   select(session_id,
#          age,
#          sex,
#          choosetobe,
#          feel,
#          hopetolive,
#          othersthink,
#          atoma,  # A person moves from being an adult to middle-aged at what age?
#          ctoya,  # A person moves from being a child to being a young adult at what age?
#          matoo,  # A person moves from being middle-aged to being old at what age?
#          yatoa) %>%  # A person moves from being a young adult to being an adult at what age?
#   filter(!is.na(age)) %>%
#   filter(sex == "m" | sex == "f")
# 
# choosetobe_df <-
#   trimmed_df %>%
#   filter(!is.na(choosetobe)) %>%
#   select(age, choosetobe, sex) %>%
#   mutate(age = as.integer(age),
#          choosetobe = as.integer(choosetobe))

# save data to disk
#save(choosetobe_df, file = "preferred age data.rdata")

# load rdata file
load(file = "data/preferred age data.rdata")
```

# Research question
"How does age related to the age people would choose to be?"

Descriptive stats

```{r choose to be - descriptives}
descriptives <- describe(choosetobe_df) %>% select(n, mean, sd)
descriptives
```

## Plots
### Age distribution

```{r }
mean_age <- descriptives["age","mean"]

ggplot(data = choosetobe_df) +
  geom_density(aes(x = age), adjust = 4, colour = "red", fill = "red", alpha = 0.2) +
  geom_vline(xintercept = mean_age, colour = "red") +
  theme_minimal()
```

### Preferred age distribution

```{r}
mean_agechoosetobe <- descriptives["choosetobe","mean"]

ggplot(data = choosetobe_df) +
  geom_density(aes(x = choosetobe), adjust = 4, colour = "blue", fill = "blue", alpha = 0.2) +
  geom_vline(xintercept = mean_agechoosetobe, colour = "blue") +
  theme_minimal()
```

Choose to be split by gender

```{r }
ggplot(data = choosetobe_df) +
  geom_density(aes(x = choosetobe, group = sex, fill = sex, colour = sex), adjust = 4, alpha = 0.2) +
  theme_minimal()
```

### Comparison

```{r}
ggplot(data = choosetobe_df) +
  geom_density(aes(x = age), adjust = 4, colour = "red", fill = "red", alpha = 0.2) +
  geom_vline(xintercept = mean_age, colour = "red") +
  geom_density(aes(x = choosetobe), adjust = 4, colour = "blue", fill = "blue", alpha = 0.2) +
  geom_vline(xintercept = mean_agechoosetobe, colour = "blue") +
  theme_minimal()
```

## Comparing distributions using t tests

```{r}
attach(choosetobe_df)

t.test(age, choosetobe, 
       alternative = "two.sided",
       paired = TRUE)
```

## Comparing 

### Plot raw data using scatterplot

```{r}
ggplot(data = choosetobe_df) +
  geom_point(aes(x = age, y = choosetobe), colour = "darkgreen", alpha = 0.1) +
  theme_minimal()
```

```{r}
ggplot(data = choosetobe_df) +
  geom_point(aes(x = age, y = choosetobe), colour = "darkgreen", alpha = 0.1) +
  geom_smooth(aes(x = age, y = choosetobe), method = "lm", colour = "darkorange") +
  theme_minimal()
```


```{r}
require(schoRsch)

choosetobe_trimmed_outlier_df <-
  choosetobe_df %>%
  schoRsch::outlier(dv = "age", 
                    res.name = "age_outlier",
                    todo = "nothing", 
                    upper.z = 2.5, 
                    lower.z = -2.5) %>%
  schoRsch::outlier(dv = "choosetobe",                   # not producing consisent results!
                    res.name = "choosetobe_outlier",
                    todo = "nothing", 
                    upper.z = 2.5, 
                    lower.z = -2.5) %>%  
  mutate(exclude = age_outlier + choosetobe_outlier) %>%
  filter(exclude == 0)

ggplot(data = choosetobe_trimmed_outlier_df) +
  geom_point(aes(x = age, y = choosetobe, group = sex, fill = sex, colour = sex), alpha = 0.1) +
  geom_smooth(aes(x = age, y = choosetobe, group = sex, colour = sex), method = "lm") +
  theme_minimal()
```


```{r}
# linear regression
model_1 <- lm(choosetobe ~ age,
              data = choosetobe_trimmed_outlier_df)
summary(model_1)

# library(apaTables)
# apa.reg.table(model_1, filename="model_1.doc", table.number=1)
```

Half their age plus 10

NB it is equally correct to say that participants want to be 2 years younger than they are on average, and that participants want to be half their age plus 10. 
