---
title: "How old is old? Teaching regression with open data"
author: "Ian Hussey^[Ghent University. Email: ian.hussey@ugent.be]"
output:
  html_document:
    code_folding: hide
    highlight: haddock
    number_sections: yes
    theme: flatly
    toc: yes
    toc_float: yes
---

# Research questions

1. What age would people chose to be if given the choice?
2. How old is old? 

## Open data

This data was collected on [Project Implicit](https://implicit.harvard.edu/) and is freely available under a [CCO license](https://creativecommons.org/publicdomain/zero/1.0/) on the [Open Science Framework](https://osf.io/pbzcu/). 

Specifically, I took the data from the Age related Implicit Association Test collected in 2015. 

```{r get and clean data, message=FALSE, warning=FALSE}

# dependencies ------------------------------------------------------------

library(foreign)
library(tidyverse)
library(psych)
library(knitr)
library(broom)

# rounds all numerics in a df
round_df <- function(df, digits) {
  nums <- vapply(df, is.numeric, FUN.VALUE = logical(1))
  df[,nums] <- round(df[,nums], digits = digits)
  (df)
}


## data processing of the raw data is comma'd out
# get data ----------------------------------------------------------------

# data from project implicit demo datasets for Age IAT 2015 data https://osf.io/pbzcu/

# convert spss to rdata
#df <- read.spss("Age IAT.public.2015.sav", to.data.frame=TRUE)  # this is the specific spss file you'd need to download to recreate these analyses

# tidy data ---------------------------------------------------------------

# trimmed_df <-
#   df %>%
#   select(session_id,
#          age,
#          sex,
#          choosetobe,
#          feel,
#          hopetolive,
#          othersthink,
#          atoma,  # A person moves from being an adult to middle-aged at what age?
#          ctoya,  # A person moves from being a child to being a young adult at what age?
#          matoo,  # A person moves from being middle-aged to being old at what age?
#          yatoa) %>%  # A person moves from being a young adult to being an adult at what age?
#   filter(!is.na(age)) %>%
#   filter(sex == "m" | sex == "f")
# 
# choosetobe_df <-
#   trimmed_df %>%
#   filter(!is.na(choosetobe)) %>%
#   select(age, choosetobe, sex) %>%
#   mutate(age = as.integer(age),
#          choosetobe = as.integer(choosetobe))

# save data to disk
#save(choosetobe_df, file = "preferred age data.rdata")

# load rdata file
load(file = "data/preferred age data.rdata")

load(file = "data/Age IAT.public.2015.rdata")

trimmed_df <-
  df %>%
  select(session_id,
         age,
         sex,
         choosetobe,
         feel,
         hopetolive,
         othersthink,
         atoma,  # A person moves from being an adult to middle-aged at what age?
         ctoya,  # A person moves from being a child to being a young adult at what age?
         matoo,  # A person moves from being middle-aged to being old at what age?
         yatoa) %>%  # A person moves from being a young adult to being an adult at what age?
  filter(!is.na(age)) %>%
  filter(sex == "m" | sex == "f")

choosetobe_data <-
  trimmed_df %>%
  filter(!is.na(choosetobe)) %>%
  select(age, choosetobe, sex) %>%
  mutate(age = as.integer(age),
         choosetobe = as.integer(choosetobe))
```

This gives us a lot of data to answer our questions:

```{r choose to be - descriptives, message=FALSE, warning=FALSE}
descriptives <- choosetobe_df %>%
  select(age) %>%
  describe() %>% 
  select(n, mean, sd)

descriptives %>%
  round_df(digits = 2) %>%
  kable()
```

# RQ1: What age would people chose to be if given the choice?

## Plot the data

I've added the usual least squared error regression line to allow us to make loose predictions from the plot.

```{r, message=FALSE, warning=FALSE}
ggplot(data = choosetobe_df) +
  geom_point(aes(x = age, y = choosetobe), colour = "darkgreen", alpha = 0.1) +
  geom_smooth(aes(x = age, y = choosetobe), method = "lm", colour = "darkorange") +
  theme_minimal()
```

## Linear regression

A linear regression can be used to make more specific predictions.

```{r, message=FALSE, warning=FALSE}
model_1 <- lm(choosetobe ~ age,
              data = choosetobe_df)
summary(model_1)
```

Conclusion: People want to be roughly half their age plus 10: current age * age estimate (0.52) + intercept (10.47)

# RQ2: How old is old?

```{r, message=FALSE, warning=FALSE}
# reshape age transitions data
transitions <- 
  trimmed_df %>%
  rename(`adult to middle` = atoma,  # A person moves from being an adult to middle-aged at what age?
         `child to ya` = ctoya,  # A person moves from being a child to being a young adult at what age?
         `middle to old` = matoo,  # A person moves from being middle-aged to being old at what age?
         `ya to adult` = yatoa) %>%
  gather(transition_type, transition_age, c(`adult to middle`, `child to ya`, `middle to old`, `ya to adult`)) %>%
  mutate(transition_type = as.factor(transition_type)) %>%
  filter(!is.na(transition_age), !is.na(transition_type))
```

## Plot the data

```{r, message=FALSE, warning=FALSE}
ggplot(transitions, aes(x = transition_age, colour = transition_type, fill = transition_type)) +
  geom_density(alpha = 0.5, adjust = 5) +  # try changing the kernal smoothing width via adjust.
  theme_minimal()
```

Rough preictions can be made from the graph. Interestingly, there seems to be more consensus (narrower distributions) about younger age transitions than old ones. 

## Linear regression

```{r, message=FALSE, warning=FALSE}
model_2 <- lm(formula = transition_age ~ transition_type,
              data = transitions)
summary(model_2)
```

Childhood ends and young adulthood begins at 16: intercept (40) + transition_type child to young adult (-24)

Young adulthood ends and adulthood begins at 22: intercept (40) + transition_type young adult to adult (-18)

Adulthood ends and middle age beings at 40: intercept (40)

Middle age ends and old age beings at 63: intercept (40) + transition_typemiddle to old (23)

# Going deeper

What if beliefs about age categories change as we age? More formally, is current age a moderator of beliefs about age categories? 

## Plot the data again

```{r, message=FALSE, warning=FALSE}
ggplot(transitions, 
       aes(x = age, y = transition_age, colour = transition_type, fill = transition_type)) +
  geom_smooth(method = "lm") +
  theme_minimal()
```

Beliefs about childhood and young adulthood are relatively stable over time. However, beliefs about middle age and old age increase with age. 

## Linear regression with interactions (moderation analysis)

```{r, message=FALSE, warning=FALSE}
model_3 <- lm(formula = transition_age ~ age * transition_type,
              data = transitions)
summary(model_3)
```

Specific predictions are harder to work out from the table here, but are consistent with inferences made from the plot.

